---
title: "Report on Lake Ontario Microbes"
author: "Christine Yang"
date: "2026-02-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

First, we need to ensure we have access to the functions that we need for our analysis.

```{r load-packages}
library(tidyverse)
```

# Goal Of This Report

Here, I create an analysis of environmental and microbial data from Lake Ontario. The goal is to determine how environmental variables like temperature affect the abundance of different Phyla of bacteria.

# Read In The Data

We need two files: 1) The Taxon File (Cleaned Up); 2) The Sample Data

## 1) Taxonomy Data

```{r load-data-taxon}
# Load The Taxon Abundance Data
taxon_dirty_df <-
  read_csv("data/taxon_abundance.csv", skip = 2) %>%
  select(-...10) %>%
  rename(sequencer = ...9)

# Look At Top Of Data-frame
head(taxon_dirty_df, 6)
# View(taxon_dirty_df)

# Clean Up Data By Removing Columns
taxon_clean_df <-
  taxon_dirty_df %>%
  select(sample_id:Cyanobacteria)

# How Many Columns? Intuition Check!
ncol(taxon_clean_df)
ncol(taxon_dirty_df)
```

Let's Load Second Data-set!

## 2) Enviromental Data

```{r load-data-samples}
sample_data_df <-
  read_csv("data/sample_data.csv")

# Quick Look
head(sample_data_df)
```

# Reshaping The Data

Create long-formatted data from wide-formatted data.

```{r reshape-data}
taxon_long_df <-
  taxon_clean_df %>%
    pivot_longer(cols = Proteobacteria:Cyanobacteria,
               names_to = "Phylum", values_to = "Abundance")

# Intuition Check
nrow(taxon_long_df)
nrow(taxon_clean_df)
```

## Plot Taxon Data

```{r plot-taxa}
taxon_long_df %>%
  ggplot(aes(x = sample_id, y = Abundance, fill = Phylum)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) 
```



